---
title: "Models with Zeros: (Multivariate) Tobit and Hurdle Models Using Stan"
author: "Hernan Bruno"
date: "EMAC 2018"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Tobit 

This script generates data and estimates a tobit model with 
heterogeneiety. In this particular example, we assume that households
may or may not purchase from a category

```{r}
library(dplyr)
library(rstan)
library(ggplot2)
```


Initialize

```{r}
set.seed(1408)
N_hh    = 200             # number of households
N_purch = 20              # number of purchases per household
alpha_0 = 1               # intercept
sigma_alpha = .5          # variability across households
sigma_y = 1
beta    = -1
```

Create the simulated dataset.
```{r}
alpha_i = rnorm(N_hh, 0, sigma_alpha)
x       = runif(N_hh*N_purch)
df_alpha = data.frame(hh_id = 1:N_hh,
                      alpha_i = alpha_i)

sym_data = data.frame(hh_id = rep(df_alpha$hh_id, each=N_purch))
sym_data = left_join(sym_data, df_alpha)
sym_data$x = x
sym_data = mutate(sym_data,
                  ystar = alpha_0 +
                          beta*x +
                          alpha_i +
                          rnorm(n(), 0, sigma_y),
                  y = pmax(0, ystar),
                  y_cen = ifelse(y == 0, 1, 0))

data_stan = list(N = nrow(sym_data),
                 N_hh = N_hh,
                 hh_id = sym_data$hh_id,
                 y = sym_data$y,
                 y_cen = sym_data$y_cen,
                 n_cen = sum(sym_data$y_cen))
```

```{r}
stan_code_tobit="data {
  int<lower=1> N;
  int<lower=1> N_hh;
  int<lower=1> hh_id[N];           // id of individuals
  real y[N];                       // Censored data (assumed at zero)
  int<lower=0, upper=1> y_cen[N];  // is left-censored : +1, 0
  real x[N]; // covariates
  int<lower=1> n_cen; 
}
parameters{
  real<upper=0> y_miss[n_cen];
  real alpha_0;
  real alpha_i[N_hh];
  real beta;
  real<lower = 0> sigma_y;
  real<lower = 0> sigma_alpha;
}

model {
  real y_c;
  int pos = 1;

  alpha_i  ~ normal(0 , sigma_alpha);

  for (n in 1:N){
    if (y_cen[n] == 0){
      y_c = y[n];
    } else {
      y_c = y_miss[pos];
      pos = pos + 1;
    }
    y_c ~ normal(alpha_0 + alpha_i[hh_id[n]] + beta*x[n], sigma_y);
  }
}"
```

```{r}
fit_tobit = stan(model_code = stan_code_tobit,
                  data = data_stan,
                  chains = 4, iter = 2000, thin=2)

print(fit_tobit, c("alpha_0", "sigma_alpha", "beta"))
```

```{r}
ext_tobit = extract(fit_tobit)
par(mfrow=c(3, 1), mar = c(3, 2, 3, 2))
plot(density(ext_tobit$alpha_0),
    main = "alpha", xlab = "")
abline(v=1, col = "blue")
abline(v=mean(ext_tobit$alpha_0), col="red", lty = 2)

plot(density(ext_tobit$sigma_alpha),
     main = "sigma_alpha", xlab = "")
abline(v=.5, col="blue")
abline(v=mean(ext_tobit$sigma_alpha), col="red", lty = 2)

plot(density(ext_tobit$beta),
    main = "beta", xlab = "")
abline(v=-1, col="blue")
abline(v=mean(ext_tobit$beta), col="red", lty=2)

daplot_tobit = data.frame(df_alpha,
                       q025 = apply(ext_tobit$alpha_i, 2, quantile, probs =.025),
                       q975 = apply(ext_tobit$alpha_i, 2, quantile, probs =.975),
                       m_alpha = apply(ext_tobit$alpha_i, 2, mean))

plot1 = ggplot(data= daplot_tobit,
               aes(x= alpha_i, ymin=q025, ymax=q975))+
       geom_linerange(colour="grey43") +
       geom_abline(intercept = 0, slope = 1, colour = "blue") +
       geom_point(data = daplot_tobit,
                  aes(x = alpha_i, y = m_alpha),
                  colour = "red", size = .75) +
       xlab("True individual alpha") +
       ylab("Estimated indivdiual alpha")
  
plot1
```

# Hurdle

    This script generates data and estimates a hurdle model. It is a
simplified model of a two-stage price change, where as salesperson first
decides whether to change the price and if yes, by how much.
A more advanced model would have a multinomial first stage, with prices
going up, down, or staying the same. Such a model would be a simple 
extension of this one.

    Particularly important is the ability of stan to introduced multivariate
heterogeneity at the salesperson level. We assume that each salesperson 
has her/his own tendency to change prices. They also might have their own
tendency to lower or increase prices. In this particular example, we assume
that these two salesperson characteristics are negatively correlated. 
That is to say, those salespeople more likely to change prices, are also
those more likely to lower prices. This need not be the case in real life.
We also assume there are covariates that can influence these decisions.

```{r}
library(mvtnorm)
library(dplyr)
library(rstan)
```

```{r}
# Initialize
set.seed(1408)
N_speople = 200  # How many individual decision makers
N_trans = 50     # How many transactions
Nobs = N_speople * N_trans
```

   Suggestion: explore situations with many salespeople and few transactions
and compare it to a situation with few salespeople with lots of transactions.

The model has a hurdle where 
```
prob_change  = logit^{-1}( alpha_i + alpha_1 x1 + alpha_2 x3)
and a quantity equation 
pricechange  =  beta_i + beta_1 x2 + beta_2 x3)
```

```{r}

alpha_1 = .2
alpha_2 = .3
beta_1  = -.3
beta_2  = .4
```

    The individual intercepts for each step have their own standard 
deviation and are correlated by rho. Here we define the covaraiance 
matrix to simulate the data.

```{r}

s_alpha  = 1
s_beta   = .5
rho       = -.7
```

create a covariance matrix

```{r}
Sigma = matrix(ncol=2, nrow=2)
Sigma[1,1] = s_alpha^2
Sigma[2,2] = s_beta^2
Sigma[1,2] = s_alpha * s_beta * rho
Sigma[2,1] = Sigma[1,2]
```

Draw a value of alpha and beta for each salesperson.
The mean for alpha_i is set at -1 and the for beta_i at .5
```{r}
temp = rmvnorm(N_speople, c(-1, .5), Sigma)
df_sp = data.frame(spid = 1:N_speople,
                   alpha_s= temp[, 1],
                   beta_s = temp[, 2])

sym_data = data.frame(spid = rep(df_sp$spid, each = N_trans))

sym_data = left_join(sym_data, df_sp) %>%
           mutate(x_1 = runif(n()),
                  x_2 = runif(n()),
                  x_3 = runif(n()),
                  lo_change  = alpha_s + alpha_1*x_1 + alpha_2*x_3,
                  mean_change = beta_s + x_2*beta_1 + beta_2*x_3,
                  prob_change = exp(lo_change) / 
                                (1 + exp(lo_change)),
                  price_change = rbinom(n(), 1, prob_change),
                  new_price = rnorm(n(), mean_change, 1))
```

```{r}
stan_code ="
data {
  int<lower=0> N;             // observations
  int<lower=0> S;             // number of salespeople
  int<lower=0> spid[N];       // identifier of salesperson      
  real x1[N];                 //   Covariates, notice that
  real x2[N];                 // that these could be introduced
  real x3[N];                 // as a matrix.
  int <lower=0, upper=1> Pchan[N];   // indicator
  real pricechange[N];       // how much did the price change
}

parameters{
  corr_matrix[2] Omega;
  vector<lower=0>[2] tau;
  vector[2] mu_ab;
  vector[2] ab[S];
  real alpha_1;             //   The alpha's could be defined as 
  real alpha_2;             // vectors. And the beta also. 
  real beta_1;
  real beta_2;
  real <lower=0> sigma_price;
}

model{
  real lo_change;           // lo stands for log-odds
  real mean_change; 
  tau ~ cauchy(0,2.5);
  Omega ~ lkj_corr(1);
  for (s in 1:S){
    ab[s] ~ multi_normal(mu_ab, quad_form_diag(Omega, tau));
  }

  for (n in 1:N){
    mean_change = ab[spid[n], 2] + x2[n]*beta_1 +x3[n]*beta_2;
    lo_change = ab[spid[n], 1] + x1[n]*alpha_1 + x3[n]*alpha_2;
    if(Pchan[n] > 0){
      pricechange[n] ~ normal(mean_change, sigma_price);
    }
    Pchan[n] ~ bernoulli_logit(lo_change);
  }
}
"
da_stan = list(N = Nobs,
               S = N_speople,
               spid = sym_data$spid,
               x1 = sym_data$x_1,
               x2 = sym_data$x_2,
               x3 = sym_data$x_3,
               Pchan = sym_data$price_change,
               pricechange = sym_data$new_price)

fit_2stage = stan(model_code = stan_code, data = da_stan, chains = 2, iter = 1000, thin=2)

print(fit_2stage, c("tau", "Omega", "mu_ab", "alpha_1", "alpha_2", "beta_1", "beta_2",  "sigma_price"))
```

To print the table into LaTeX
```{r}
library(xtable)
xtable(print(fit_2stage, c("tau", "Omega", "mu_ab", "alpha_1", "alpha_2", "beta_1", "beta_2",  "sigma_price")))
```

